Program Equivalence : A Relational Separation Logic interactive prover implemented in Maude

### 2018-06-27 paper/paper_after_guide.tex
made the prover flow figure be placed relatively ok
second proof-reading done

\tableofcontents
\pagebreak
\begin{abstract}
	Lucrarea de față descrie procesul de dezvoltare a unui utilitar interactiv utilizat în analiza formală a programelor, folosindu-se de Logica Separatoare Relațională \cite{relational} și Logica Separatoare \cite{primer} \cite{SeparationLogic}, utilizând Maude \cite{maudesite} drept cadru de dezvoltare. Componenta interactiva se îmbină cu o componentă automatizată, prin care părți din demonstrație vor fi realizate automat. Lucrarea este orientata pe rezolvarea problemei prin intermediul funcționalităților oferite de Maude și va include secvențe de cod ce vor ilustra conceptele cheie legate de modelarea si aplicarea fundamentelor teoretice utilizate pe parcursul proiectului.
\end{abstract}
\section*{Introduction}
The present paper describes the development of an interactive tool for reasoning how two programs are related, based theoretical concepts from the field of formal program analysis and technologies which facilitate the implementation of those concepts. \\

@@

The tool is built as a CLI which helps with reasoning how two programs are related using Relational Separation Logic specifications. As a consequence of the dependency of Relational Separation Logic on Separation Logic, proofs about single programs using the latter are also supported by the tool. The tool has been developed with extensibility in mind, the main desired extensions being concurrent programs support and automatic proofs. \\

@@

Comparing programs or code fragments and studying their equivalence is part of every software engineer's activities when they are testing an alternative implementation for an existing solution, fixing bugs, launching new product versions, etc . Naturally, for every process completed manually there are efforts being made in order to make it more efficient, less error-prone and, in the end, automate the process all together. Once such a task is automated in software engineering, it can be included in the flow of any research or development phase of a product. An example benefiting from a formal proof of program equivalence is compiler optimization, where the optimized code needs to be equivalent to the input one . \\

This project aims to lay the foundations of a tool which facilitates formal reasoning on the way two programs are related with a focus on extensibility and automation of tasks.

@@

 Good references for Maude language include:
 \begin{itemize}
 	\item Maude Primer \cite{primer} which was written for Maude 2.0.1 but still applies to the latest versions of Maude is a good starting point for learning Maude, as it introduces the language in a more informal, friendlier way
 	\item Maude Manual \cite{manual} which presents the whole Maude system in depth, along with the mathematical foundations
 	\item All about Maude book \cite{allAboutMaude} which includes everything contained in the manual and presents some of the more relevant tools implemented in Maude .
 \end{itemize}
\section{Relational Separation Logic Interactive Prover} 
This section will include details related to the implementation and usage of the Relational Separation Logic prover and highlight a few of Maude's features which made it the a language fit for the purpose of this project.
\\

@@

This section will include snippets of Maude code, but since Maude is a very expressive and self-explanatory language, I have not included of thorough presentation of every language construct defined for the scope of this project. Ideas presented will be accompanied by some examples and key parts of the implementation but for a more detailed presentation, please refer to the code itself.
\subsection{Language Grammar and executable semantics}
The prover supports expressing programs in a simple, imperative language, commonly used throughout papers  \cite{primer} \cite{SeparationLogic} related to the subject of program verification, including the one describing Relational Separation Logic \cite{relational}, around  which this project is based. 
\subsubsection{Storage Model}
A state of in our storage model is defined by a pair consisting of a \(Store\) and a \(Heap\) .
\\
Assuming that all the variables usable in programs from the set \(\mathtt{Vars}\) and the set of positive natural numbers is denoted by \(\mathtt{PosNats}\): \\

@@

Each proof rule in the Separation Logic and Relational Separation Logic has been translated into a Maude rewrite rule. 
\\
The prover is centered around the concept of \textbf{Goals} which can be one of the three things: 
\begin{itemize}
	\item {Hoare Quadruple} - corresponding to Relational Separation Logic
	\item {Hoare Triple} - corresponding to Separation Logic
	\item {Implication between relations / assertions} 
\end{itemize} 
The proof rules are interpreted in a bottom-up manner, meaning that, applying a rule to a goal, which must match the conclusion of the used rule, will generate new goals consisting of the hypothesis: Figure \ref{fig:SLTranslation} presents how a Separation Logic rule has been translated into Maude syntax and Figure \ref{fig:RSLTranslation} presents the same process for the Relational Separation Logic counterpart of the rule.
\begin{figure}[h]
	Consequence Rule
	\medskip
	\[
	\begin{prooftree}
	\Hypo{P \Rightarrow P\prime}
	\Hypo{\{P\prime\} C \{Q\prime\}}
	\Hypo{Q \Rightarrow Q\prime}
	\Infer 3 {\{P\}\  C  \{Q\}}
	\end{prooftree}	
	\]
	Is represented in Maude syntax as : 
\begin{lstlisting}[caption=Separation Logic Consequence rule]
	rl [SeparationLogic-Consequence] : {P} C {Q} => ((P => P1) <> ({P1} C {Q1})) <> (Q1 => Q) [nonexec] .
\end{lstlisting}
	\caption{Example of a translation to Maude syntax of a Separation Logic proof rule}
	\label{fig:SLTranslation}
\end{figure}

\begin{figure}[h]
	Consequence Rule
	\[
	\begin{prooftree}
	\Hypo{R \Rightarrow R_1}
	\Hypo{\hq{R_1}{C}{C\prime}{S_1}}
	\Hypo{S \Rightarrow S_1}
	\Infer 3 {\hq{R}{C}{C\prime}{S}}
	\end{prooftree}	
	\]
	Is represented in Maude syntax as : 
	\begin{lstlisting}[caption=Relational Separation Logic Consequence rule]
	rl [Consequence] : { R } C1 -- C2  { S } => ((R => R1) <> ({R1} C1 -- C2 {S1})) <> (S1 => S) [nonexec] .
	\end{lstlisting}
	\caption{Example of a translation to Maude syntax of a Relational Separation Logic proof rule}
	\label{fig:RSLTranslation}
\end{figure}
The \texttt{<>} operator is used to separate sub-goals generated by the rule.
\\

@@

All the rules are marked with the attribute \texttt{[nonexec]} because we want them to be applied in a controlled manner, at the meta-level. Some of the rules, the examples included, are not even valid executable Maude rules, because they make use of variables before binding them (for example \texttt{P1} is used in a rule before it is bound). At the meta-level however we can manually bind values to those variables, as we will exemplify when discussing how we apply these rules to goals.
\subsubsection{Sort Hierarchy}
Much in the same manner as we modeled the language grammar, the Relational Separation Logic \cite{relational} and Separation Logic \cite{SeparationLogic} \cite{primer} have been modeled by defining sorts and declaring operators using these sorts. The main point of reference for our modeling has been \cite{relational}, for both logics. Some examples of translations from specification to code are included below.
\begin{itemize}
	\item \texttt{Assertion} is the sort used to handle Separation Logic assertions. It is a super sort of the \texttt{BExp} sort defined as part of our language.
	\item \texttt{HoareTriple} is the sort associated with the concept with the same name, in a Separation Logic context.
	\begin{lstlisting}[caption=Hoare Triple constructor operator]
op {_}_{_} : Assertion Command Assertion -> HoareTriple .\end{lstlisting}
	\item \texttt{Relation} is the sort associated with Relational Separation Logic relations. As it was the case with \texttt{Assertion}, it is a super sort of the \texttt{BExp} sort of our language grammar. Because both the \texttt{Relation} and \texttt{Assertion} are super sorts of the same sort, Maude will display a few warnings which are not affecting functionality, but in the future, creating a separate \texttt{BExp} sort for both the \texttt{Assertion} and \texttt{Relation} sorts would be a better option.
	\item \texttt{HoareQuadruple} is again associated with the concept with the same name, in the Relational Separation Logic context.
	\begin{figure}[h]
	\begin{lstlisting}[caption=Examples of Relational Separation Logic specific constructs]
--- Operator which creates a relation from two assertions
op [_;_] : Assertion Assertion -> Relation .
--- Operator which constructs a Hoare Quadruple 
op {_}_--_{_} : Relation Command Command Relation -> HoareQuadruple .\end{lstlisting}
	\end{figure}
\end{itemize}
The other operators for the Assertion and Relation languages have been defined in the same manner as the programming language operators, following their presentation in \cite{relational}.

@@

	\includegraphics[width=\linewidth]{"pictures/Prover Flow".png}

@@

	\item Axioms are modeled as terms with variables in them, allowing them to be matched with any term for which all variables in the axiom can be bound.

@@

	However, Maude does not allow for variables to appear for the first time on the right hand side of a rewrite rule which needs to be used without meta-level computations (problem present when modeling the Separation Logics as well). We had to find a work around for this problem and the solution we came up with was to write the axiom in its meta-representation, and then use the \texttt{downTerm} operator to bring it down in the meta-hierarchy (Listing \ref{axiom})

@@

We have modeled implication proof rules as rewrite rules in Maude (example in Listing \ref{implicationProofRule}), in order to make use of its \texttt{metaSearchPath} operator, which will search for a series of rewrites which transform a term into another.
\\
We used this functionality to try and solve implications 
\[
	R \Rightarrow S
\]
by searching for a series of rewrites from \texttt{R => S} to \texttt{true}. 
\\ 
\begin{lstlisting}[label=implicationProofRule,caption=Example of implication rewrite rules]
rl R:Relation * ( S:Relation \/ T:Relation ) => (R:Relation * S:Relation) \/ (R:Relation * T:Relation) .
--- Example for the usage of the replace operator
rl R * (Emp /\ X eqs Y) => replace(R, X , Y) .
rl (R:Relation => R:Relation) => true .
\end{lstlisting}
Most proof rules are of the form \(R \Leftrightarrow S\) and thus can be naturally expressed in our prover using rewrite rules. However, when the proof rules involve just a one way implication, such as \(Emp \Rightarrow Same\), if we were to model it simply as the first rule in the Listing \ref{implicationProofRule}, an implication \texttt{Same => Emp} would be considered true, because Maude would rewrite the \texttt{Emp} to \texttt{Same} and \texttt{Same => Same} evaluates to \texttt{true}. To prevent this from happening, we have constrained one-way implications by providing the entire context in which they can be rewritten :
\begin{lstlisting}
rl ((R * Emp) => S) => ((R * Same) => S) .
\end{lstlisting}

@@

\begin{lstlisting}[label=eqImplicationProving,caption=Equation using the metaSearchPath operator]
op tryImplicationProving : Goal Int -> Trace? .
eq tryImplicationProving(RR:Relation => TT:Relation, depth:Int) = metaSearchPath(upModule('IMPLICATION-PROVER, false), upTerm(RR:Relation => TT:Relation),upTerm(true), nil, '* , depth:Int, 0) . 
\end{lstlisting}
In it's current form, it cannot offer any guarantee, except for the certainty that if the implication has been proven automatically, the proof is correct with respect to the rewrite rules specified in the \textbf{implication-prover.maude} file. The approach however has an exponential running time since it searches exhaustively throughout the state space for a solution. \\

@@

In the future, one might consider using Maude internal strategies \cite{manual} to control the rewriting and searching in the implication proving process, to make it follow heuristics and thus be more effective.

\paragraph{Inductive predicates} - the prover currently supports defining inductive predicates only by modifying the underlying Maude code. We have defined a single predicate, used in the Relational Separation Logic paper \cite{relational} :
\begin{lstlisting}[caption=Predicate syntax definition]
op List : Id -> Relation . 
\end{lstlisting}
In terms of the semantics of the \texttt{List} predicate, we have added rules to the \textbf{implication-prover.maude} file, to be considered when rewriting implications.
\begin{lstlisting}[caption=Example of rule interpreting the List predicate during implication proving]
rl [ (X:Id |-> E:AExp) * ((X:Id plus 1) |-> Y:Id) ; (X:Id |-> E:AExp) * ((X:Id plus 1) |-> Y:Id) ] * List(Y:Id) => List(X) .
\end{lstlisting}
\paragraph{Cyclist}\cite{cyclist} \cite{cyclistSite} is an automatic entailment prover for Separation Logic which we planned to integrate with our prover for Separation Logic implications and eventually extend to Relational Separation Logic, since it provides a generic framework. Since Maude does not easily support external commands, the idea would be to modify Maude's C++ code and add a hook for the \texttt{[special]} operator attribute which would invoke that hook, which will call the external command for Cyclist. We have however focused on other aspects of our prover, and left this feature for the future.
\subsection{User Interface}
Since our project aims to simplify the demonstrations of Separation Logic / Relational Separation Logic formulae, using an interactive prover, the user interface plays a major role in the success of the project - the easier it is to use and understand, the more chances of success it has. This idea has been presented in \cite{primer}, when discussing key elements of tools related to program verification. \\

@@

For a more advanced user interface or to expose the prover's functionalities through a programming interface, a wrapper written in more feature-rich language through which to control the Maude environment is probably the best choice. The concept is exemplified by the Maude plug-in for Eclipse, \cite{maudeEclipsePLugin} which enables writing of Maude code inside the Eclipse IDE, and the plug-in communicates with the underlying Maude environment. The project also includes a plug-in which allows any Java program to communicate with Maude.
\section*{Conclusions}
The development of the Relational Separation Logic prover helped us get a deeper understating of it, along with an introduction to other theoretical concepts used in program analysis and programming language semantics, such as Separation Logic and Rewriting Logic.
\\

@@

Maude has been interesting to work because it required an entirely different mindset than the mainstream programming languages. Once grasping the concepts around Maude which was built, it was easy to come up with simple, clever solutions to the encountered problems. \\
As it is described on it's website \cite{maudesite}, Maude is a powerful logical framework which allows modeling of other logics in it; it proved to be exactly that, enabling us to represent the logics in a natural manner, which made the implementation much less error-prone and allowed us to focus on the relevant part of this project, namely how to apply the logics instead of focusing on how to properly implement them.
\\

@@

Overall, this whole project has been an interesting introduction to both  theoretical concepts regarding program analysis and new, different technologies. The mix of the two resulted in the built prover, which could be improved in the future with the features discussed throughout the chapters to easily accommodate real-world usage.

@@

\begin{appendices}
	\section{Example of proof using the prover}
	\label{proofExample}
	This appendix contains all the necessary commands to be run in a Maude console in order to run an example demonstration using the prover.
\begin{lstlisting}
load prover-interface.maude
loop init .
(prove(
{ 
((
Same 
*
List(c) 
*
[ y |-> x0 ; yp |-> x0 ]
) 
*
(y eqs yp /\ Emp))
*
(c eqs cp /\ Emp)
}
pass ; 
while (! (c eqs null))
do 
x := [y] ; 
[c] := 0 minus x ;
c := [c plus 1]
od
--
xp := [yp] ;
while (! (cp eqs null))
do 
[cp] := 0 minus xp ;
cp := [cp plus 1] 
od
{
(Same
*
(y eqs yp /\ Emp))
* 
(c eqs cp /\ Emp)
}
)
)
(applySequencing ( Same * List(c) * [ y |-> x0 ; yp |-> x0 ] * (y eqs yp /\ Emp) * (c eqs cp /\ Emp) * (xp eqs x0 /\ Emp) ) )
(applyFrameRule (Same * List(c) * (y eqs yp /\ Emp) * (c eqs cp /\ Emp)) )
(
applyConsequence
( [ (y |-> x0) ; (yp |-> x0) ] )
--
( [ (y |-> x0) ; ((yp |-> x0) * (xp eqs x0 /\ emp)) ] )
)
(applyEmbeddingRule)
(
applyConsequence
( Same * List(c) * (y eqs yp /\ Emp) * (xp eqs x0 /\ Emp) * (c eqs cp /\ Emp) * [y |-> x0 ; yp |-> x0] )
--
( (Same * List(c) * (y eqs yp /\ Emp) * (xp eqs x0 /\ Emp) * (c eqs cp /\ Emp) * [y |-> x0 ; yp |-> x0]) /\ !(!(c eqs null)) ) --- R si not B
)
(applyLoop)
(applyConsequence
([(c0 |-> -) * ((c0 plus 1) |-> c1) * (y |-> x0) * (c eqs c0 /\ emp) 
; (c0 |-> -) * ((c0 plus 1) |-> c1) * (xp eqs x0 /\ emp ) * (cp eqs c0 /\ emp) ] 
* List(c1) * Same * (y eqs yp /\ Emp) * [emp ; yp |-> x0])
--
([(c0 |-> (0 minus x0)) * ((c0 plus 1) |-> c1) * (y |-> x0) * (c eqs c1 /\ emp) 
; (c0 |-> (0 minus x0)) * ((c0 plus 1) |-> c1) * (xp eqs x0 /\ emp ) * (cp eqs c1 /\ emp) ]
* List(c1) * Same * (y eqs yp /\ Emp) * [emp ; yp |-> x0])
)
(ok) 
(applyFrameRule (List(c1) * Same * (y eqs yp /\ Emp) * [emp ; yp |-> x0]))
(applyEmbeddingRule)
(SLapplySequencing ((c0 |-> (0 minus x0)) * ((c0 plus 1) |-> c1) * (y |-> x0) * (emp /\ c eqs c0)))
(SLapplySequencing ((c0 |-> -) * ((c0 plus 1) |-> c1) * (y |-> x0) * (emp /\ c eqs c0) * (emp /\ x eqs x0)))
(SLapplyFrameRule ((c0 |-> -) * ((c0 plus 1) |-> c1) * (emp /\ c eqs c0)))
(SLapplyFrameRule ( (y |-> x0) * ((c0 plus 1)|-> c1) ) )
(SLapplyFrameRule ( (y |-> x0) *  (c0 |->(0 minus x0)) ) )
(SLapplySequencing ((c0 |-> (0 minus x0)) * ((c0 plus 1) |-> c1) * (xp eqs x0 /\ emp) * (emp /\ cp eqs c0)))
(SLapplyFrameRule ( (c0 plus 1)|-> c1 ))
(SLapplyFrameRule  ((emp /\ (xp eqs x0) ) * (c0 |->(0 minus x0))) )
(ok)
(auto 9)
\end{lstlisting}
\end{appendices}

>>>>

\tableofcontents
\pagebreak
\begin{abstract}
	Lucrarea de față descrie procesul de dezvoltare al unui utilitar interactiv utilizat în analiza formală a programelor, folosindu-se de Logica Separatoare Relațională \cite{relational} și Logica Separatoare \cite{primer} \cite{SeparationLogic}, utilizând Maude \cite{maudesite} drept cadru de dezvoltare. Componenta interactivă se îmbină cu o componentă automatizată, prin care părți din demonstrație vor fi realizate automat. Lucrarea este orientata pe rezolvarea problemei prin intermediul funcționalităților oferite de Maude și va include secvențe de cod ce vor ilustra conceptele cheie legate de modelarea si aplicarea fundamentelor teoretice utilizate pe parcursul proiectului.
\end{abstract}
\section*{Introduction}
The present paper describes the development of an interactive tool for reasoning how two programs are related, based theoretical concepts from the field of formal program analysis and technologies which facilitate the implementation of those concepts. \\

@@

The tool is built as a CLI which helps with reasoning about how two programs are related using Relational Separation Logic specifications. As a consequence of the dependency of Relational Separation Logic on Separation Logic, proofs about single programs using the latter are also supported by the tool. The tool has been developed with extensibility in mind, the main desired extensions being concurrent programs support and automatic proofs. \\

@@

Comparing programs or code fragments and studying their equivalence is part of every software engineer's activities when they are testing an alternative implementation for an existing solution, fixing bugs, launching new product versions, etc . Naturally, for every process completed manually there are efforts being made in order to make it more efficient, less error-prone and, in the end, automate the process all together. Once such a task is automated in software engineering, it can be included in the flow of any research or development phase of a product. An example benefiting from a formal proof of program equivalence is compiler optimization, where the optimized code needs to be equivalent to the input one \cite{relational} . \\

This project aims to lay the foundations of a tool which facilitates formal reasoning about the way two programs are related with a focus on extensibility and automation of tasks.

@@

 Good references for Maude language include:
 \begin{itemize}
 	\item Maude Primer \cite{primer} which was written for Maude 2.0.1 but still applies to the latest versions of Maude is a good starting point for learning Maude, as it introduces the language in a more informal, friendlier way
 	\item Maude Manual \cite{manual} which presents the whole Maude system in depth, along with the mathematical foundations
 	\item All about Maude book \cite{allAboutMaude} which includes everything contained in the manual and presents some of the more relevant tools implemented in Maude .
 \end{itemize}
\section{Relational Separation Logic Interactive Prover} 
This section will include details related to the implementation and usage of the Relational Separation Logic prover and highlight a few of Maude's features which made it a language fit for the purpose of this project.
\\

@@

This section will include snippets of Maude code, but since Maude is a very expressive and self-explanatory language, I have not included of thorough presentation of every language construct defined for the scope of this project. Ideas presented will be accompanied by some examples and key parts of the implementation but for a more detailed presentation, please refer to the code itself.
\subsection{Language}
The prover supports expressing programs in a simple, imperative language, commonly used throughout papers  \cite{primer} \cite{SeparationLogic} related to the subject of program verification, including the one describing Relational Separation Logic \cite{relational}, around  which this project is based. 
\subsubsection{Storage Model}
A state of in our storage model is defined by a pair consisting of a \(Store\) and a \(Heap\) .
\\
Assuming that all the variables usable in programs from the set \(\mathtt{Vars}\) and the set of positive natural numbers is denoted by \(\mathtt{PosNats}\): \\

@@

Each proof rule in the Separation Logic and Relational Separation Logic has been translated into a Maude rewrite rule. 
\\
The prover is centered around the concept of \textbf{Goals} which can be one of the three things: 
\begin{itemize}
	\item {Hoare Quadruple} - corresponding to Relational Separation Logic
	\item {Hoare Triple} - corresponding to Separation Logic
	\item {Implication between relations / assertions} 
\end{itemize} 
The proof rules are interpreted in a bottom-up manner, meaning that, applying a rule to a goal, which must match the conclusion of the used rule, will generate new goals consisting of the hypothesis: Figure \ref{fig:SLTranslation} presents how a Separation Logic rule has been translated into Maude syntax and Figure \ref{fig:RSLTranslation} presents the same process for the Relational Separation Logic counterpart of the rule.
\begin{figure}[h]
	Consequence Rule
	\medskip
	\[
	\begin{prooftree}
	\Hypo{P \Rightarrow P\prime}
	\Hypo{\{P\prime\} C \{Q\prime\}}
	\Hypo{Q\prime \Rightarrow Q}
	\Infer 3 {\{P\}\  C  \{Q\}}
	\end{prooftree}	
	\]
	Is represented in Maude syntax as : 
\begin{lstlisting}[caption=Separation Logic Consequence rule]
	rl [SeparationLogic-Consequence] : {P} C {Q} => ((P => P1) <> ({P1} C {Q1})) <> (Q1 => Q) [nonexec] .
\end{lstlisting}
	\caption{Example of a translation to Maude syntax of a Separation Logic proof rule}
	\label{fig:SLTranslation}
\end{figure}

\begin{figure}[h]
	Consequence Rule
	\[
	\begin{prooftree}
	\Hypo{R \Rightarrow R_1}
	\Hypo{\hq{R_1}{C}{C\prime}{S_1}}
	\Hypo{S_1 \Rightarrow S}
	\Infer 3 {\hq{R}{C}{C\prime}{S}}
	\end{prooftree}	
	\]
	Is represented in Maude syntax as : 
	\begin{lstlisting}[caption=Relational Separation Logic Consequence rule]
	rl [Consequence] : { R } C1 -- C2  { S } => ((R => R1) <> ({R1} C1 -- C2 {S1})) <> (S1 => S) [nonexec] .
	\end{lstlisting}
	\caption{Example of a translation to Maude syntax of a Relational Separation Logic proof rule}
	\label{fig:RSLTranslation}
\end{figure}
The \texttt{<>} operator is used to separate sub-goals generated by the rule.
\\

@@

All the rules are marked with the attribute \texttt{[nonexec]} because we want them to be applied in a controlled manner, at the meta-level. Some of the rules, the examples included, are not even valid executable Maude rules, because they make use of variables before binding them (for example \texttt{P1} is used in a rule before it is bound). At the meta-level however we can manually bind values to those variables, as we will exemplify when discussing how we apply these rules to goals.
\subsubsection{Sort Hierarchy}
Much in the same manner as we modelled the language grammar, the Relational Separation Logic \cite{relational} and Separation Logic \cite{SeparationLogic} \cite{primer} have been modeled by defining sorts and declaring operators using these sorts. The main point of reference for our modelling has been \cite{relational}, for both logics. Some examples of translations from specification to code are included below.
\begin{itemize}
	\item \texttt{Assertion} is the sort used to handle Separation Logic assertions. It is a super sort of the \texttt{BExp} sort defined as part of our language.
	\item \texttt{HoareTriple} is the sort associated with the concept with the same name, in a Separation Logic context.
	\begin{lstlisting}[caption=Hoare Triple constructor operator]
op {_}_{_} : Assertion Command Assertion -> HoareTriple .\end{lstlisting}
	\item \texttt{Relation} is the sort associated with Relational Separation Logic relations. As it was the case with \texttt{Assertion}, it is a super sort of the \texttt{BExp} sort of our language grammar. Because both the \texttt{Relation} and \texttt{Assertion} are super sorts of the same sort, Maude will display a few warnings which are not affecting functionality, but in the future, creating a separate \texttt{BExp} sort for both the \texttt{Assertion} and \texttt{Relation} sorts would be a better option.
	\item \texttt{HoareQuadruple} is again associated with the concept with the same name, in the Relational Separation Logic context.
	\begin{figure}[h]
	\begin{lstlisting}[caption=Examples of Relational Separation Logic specific constructs]
--- Operator which creates a relation from two assertions
op [_;_] : Assertion Assertion -> Relation .
--- Operator which constructs a Hoare Quadruple 
op {_}_--_{_} : Relation Command Command Relation -> HoareQuadruple .\end{lstlisting}
	\end{figure}
\end{itemize}
The other operators for the Assertion and Relation languages have been defined in the same manner as the programming language operators, following their presentation in \cite{relational}.

@@

\begin{figure}[h!]
	\includegraphics[width=\textwidth]{"pictures/Prover Flow".png}

@@

	\item Axioms are modelled as terms with variables in them, allowing them to be matched with any term for which all variables in the axiom can be bound.

@@

	However, Maude does not allow for variables to appear for the first time on the right hand side of a rewrite rule which needs to be used without meta-level computations (problem present when modelling the Separation Logics as well). We had to find a work around for this problem and the solution we came up with was to write the axiom in its meta-representation, and then use the \texttt{downTerm} operator to bring it down in the meta-hierarchy (Listing \ref{axiom})

@@

We have modelled implication proof rules as rewrite rules in Maude (example in Listing \ref{implicationProofRule}), in order to make use of its \texttt{metaSearchPath} operator, which will search for a series of rewrites which transform a term into another.
\\
We used this functionality to try and solve implications 
\[
	R \Rightarrow S
\]
by searching for a series of rewrites from \texttt{R => S} to \texttt{true}. 
\\ 
\begin{lstlisting}[label=implicationProofRule,caption=Example of implication rewrite rules]
rl R:Relation * ( S:Relation \/ T:Relation ) => (R:Relation * S:Relation) \/ (R:Relation * T:Relation) .
--- Example for the usage of the replace operator
rl R * (Emp /\ X eqs Y) => replace(R, X , Y) .
rl (R:Relation => R:Relation) => true .
\end{lstlisting}
Most proof rules are of the form \(R \Leftrightarrow S\) and thus can be naturally expressed in our prover using rewrite rules. However, when the proof rules involve just a one way implication, such as \(Emp \Rightarrow Same\), if we were to model it simply as the first rule in the Listing \ref{implicationProofRule}, an implication, \texttt{Same => Emp} for example, could be wrongly considered true, because Maude would rewrite the \texttt{Emp} to \texttt{Same} and \texttt{Same => Same} evaluates to \texttt{true}. To prevent this from happening, we have constrained one-way implications by providing the entire context in which they can be rewritten :
\begin{lstlisting}
rl ((R * Emp) => S) => ((R * Same) => S) .
\end{lstlisting}

@@

\begin{lstlisting}[label=eqImplicationProving,caption=Equation using the metaSearchPath operator]
op tryImplicationProving : Goal Int -> Trace? .
eq tryImplicationProving(RR:Relation => TT:Relation, depth:Int) = metaSearchPath(upModule('IMPLICATION-PROVER, false), upTerm(RR:Relation => TT:Relation),upTerm(true), nil, '* , depth:Int, 0) . 
\end{lstlisting}
In its current form, the prover does not offer any guarantees, except for the certainty that if the implication has been proven automatically, the proof is correct with respect to the rewrite rules specified in the \textbf{implication-prover.maude} file. The approach however has an exponential running time since it searches exhaustively throughout the state space for a solution. \\

@@

Almost all the implications in Appendix \ref{proofExample} can be automatically proven, the exceptions being the ones for which the running time of the prover is too high, since they require a depth higher than 15 - those are manually admitted using the \texttt{ok} command.
\\

In the future, one might consider using Maude internal strategies \cite{manual} to control the rewriting and searching in the implication proving process, to make it follow heuristics and thus be more effective.
\paragraph{Inductive predicates} - the prover currently supports defining inductive predicates only by modifying the underlying Maude code. We have defined a single predicate, used in the Relational Separation Logic paper \cite{relational} :
\begin{lstlisting}[caption=Predicate syntax definition]
op List : Id -> Relation . 
\end{lstlisting}
In terms of the semantics of the \texttt{List} predicate, we have added rules to the \textbf{implication-prover.maude} file, to be considered when rewriting implications.
\begin{lstlisting}[caption=Example of rule interpreting the List predicate during implication proving]
rl [ (X:Id |-> E:AExp) * ((X:Id plus 1) |-> Y:Id) ; (X:Id |-> E:AExp) * ((X:Id plus 1) |-> Y:Id) ] * List(Y:Id) => List(X) .
\end{lstlisting}
\paragraph{Cyclist}\cite{cyclist} \cite{cyclistSite} is an automatic entailment prover for Separation Logic which we planned to integrate with our prover for Separation Logic implications and eventually extend to Relational Separation Logic, since it provides a generic framework. Since Maude does not easily support external commands, the idea would be to modify Maude's C++ code and add a hook for the \texttt{[special]} operator attribute which would invoke that hook, which will call the external command for Cyclist. We have however focused on other aspects of our prover, and left this feature for the future.
\subsection{User Interface}
Since our project aims to simplify the demonstrations of Separation Logic / Relational Separation Logic formulae, using an interactive prover, the user interface plays a major role in the success of the project - the easier it is to use and understand, the more chances of success it has. This idea has been presented in \cite{primer}, when discussing key elements of tools related to program verification. \\

@@

For a more advanced user interface or to expose the prover's functionalities through a programming interface, a wrapper written in more feature-rich language through which to control the Maude environment is probably the best choice. The concept is exemplified by the Maude plug-in for Eclipse, \cite{maudeEclipsePLugin} which enables writing of Maude code inside the Eclipse IDE, and the plug-in communicates with the underlying Maude environment. The project also includes a generic plug-in which allows any Java program to communicate with Maude.
\section*{Conclusions}
The development of the Relational Separation Logic prover helped us get a deeper understating of it, along with an introduction to other theoretical concepts used in program analysis and programming language semantics, such as Separation Logic and Rewriting Logic.
\\

@@

Maude has been interesting to work because it required an entirely different mindset than the mainstream programming languages. Once grasping the concepts around Maude which was built, it was easy to come up with simple, clever solutions to the encountered problems. \\
As it is described on it's website \cite{maudesite}, Maude is a powerful logical framework which allows modelling of other logics in it; it proved to be exactly that, enabling us to represent the logics in a natural manner, which made the implementation much less error-prone and allowed us to focus on the relevant part of this project, namely how to apply the logics instead of focusing on how to properly implement them.
\\

@@

Overall, this whole project has been an interesting introduction to both  theoretical concepts regarding program analysis and new, different technologies. The mix of the two resulted in the built prover, which could be improved in the future, both from an implementation standpoint (features discussed throughout the chapters) and by refining the theoretical concepts on which it is based (concurrency extensions for logics) to easily accommodate real-world usage.

@@

\begin{appendices}
	\section{Example of proof using the prover}
	\label{proofExample}
	This appendix contains all the necessary commands to be run in a Maude console in order to run an example demonstration using the prover.
\begin{lstlisting}
load prover-interface.maude
loop init .
(prove(
{ 
((
Same 
*
List(c) 
*
[ y |-> x0 ; yp |-> x0 ]
) 
*
(y eqs yp /\ Emp))
*
(c eqs cp /\ Emp)
}
pass ; 
while (! (c eqs null))
do 
x := [y] ; 
[c] := 0 minus x ;
c := [c plus 1]
od
--
xp := [yp] ;
while (! (cp eqs null))
do 
[cp] := 0 minus xp ;
cp := [cp plus 1] 
od
{
(Same
*
(y eqs yp /\ Emp))
* 
(c eqs cp /\ Emp)
}
)
)
(applySequencing ( Same * List(c) * [ y |-> x0 ; yp |-> x0 ] * (y eqs yp /\ Emp) * (c eqs cp /\ Emp) * (xp eqs x0 /\ Emp) ) )
(applyFrameRule (Same * List(c) * (y eqs yp /\ Emp) * (c eqs cp /\ Emp)) )
(
applyConsequence
( [ (y |-> x0) ; (yp |-> x0) ] )
--
( [ (y |-> x0) ; ((yp |-> x0) * (xp eqs x0 /\ emp)) ] )
)
(applyEmbeddingRule)
(
applyConsequence
( Same * List(c) * (y eqs yp /\ Emp) * (xp eqs x0 /\ Emp) * (c eqs cp /\ Emp) * [y |-> x0 ; yp |-> x0] )
--
( (Same * List(c) * (y eqs yp /\ Emp) * (xp eqs x0 /\ Emp) * (c eqs cp /\ Emp) * [y |-> x0 ; yp |-> x0]) /\ !(!(c eqs null)) )
)
(applyLoop)
(applyConsequence
([(c0 |-> -) * ((c0 plus 1) |-> c1) * (y |-> x0) * (c eqs c0 /\ emp) 
; (c0 |-> -) * ((c0 plus 1) |-> c1) * (xp eqs x0 /\ emp ) * (cp eqs c0 /\ emp) ] 
* List(c1) * Same * (y eqs yp /\ Emp) * [emp ; yp |-> x0])
--
([(c0 |-> (0 minus x0)) * ((c0 plus 1) |-> c1) * (y |-> x0) * (c eqs c1 /\ emp) 
; (c0 |-> (0 minus x0)) * ((c0 plus 1) |-> c1) * (xp eqs x0 /\ emp ) * (cp eqs c1 /\ emp) ]
* List(c1) * Same * (y eqs yp /\ Emp) * [emp ; yp |-> x0])
)
(ok) 
(applyFrameRule (List(c1) * Same * (y eqs yp /\ Emp) * [emp ; yp |-> x0]))
(applyEmbeddingRule)
(SLapplySequencing ((c0 |-> (0 minus x0)) * ((c0 plus 1) |-> c1) * (y |-> x0) * (emp /\ c eqs c0)))
(SLapplySequencing ((c0 |-> -) * ((c0 plus 1) |-> c1) * (y |-> x0) * (emp /\ c eqs c0) * (emp /\ x eqs x0)))
(SLapplyFrameRule ((c0 |-> -) * ((c0 plus 1) |-> c1) * (emp /\ c eqs c0)))
(SLapplyFrameRule ( (y |-> x0) * ((c0 plus 1)|-> c1) ) )
(SLapplyFrameRule ( (y |-> x0) *  (c0 |->(0 minus x0)) ) )
(SLapplySequencing ((c0 |-> (0 minus x0)) * ((c0 plus 1) |-> c1) * (xp eqs x0 /\ emp) * (emp /\ cp eqs c0)))
(SLapplyFrameRule ( (c0 plus 1)|-> c1 ))
(SLapplyFrameRule  ((emp /\ (xp eqs x0) ) * (c0 |->(0 minus x0))) )
(ok)
(auto 9)
\end{lstlisting}
\end{appendices}
